---
title: "ek_counterfactuals"
author: "Aleksander Hycnar"
date: "2026-09-22"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(here)
library(tidyverse)
```
## Question 1

### Write down the model

\begin{equation}
\sum_{i} X^0_{ni} = \text{total spending of } n
\label{eq:rowsum}
\end{equation}

\begin{equation}
\sum_{d} X^0_{di} = \text{total sales, hence income, of } i = w^0_i L^0_i
\label{eq:colsum}
\end{equation}

\begin{equation}
F_i(z) = \exp\left(-T_i z^{-\theta}\right)
\label{eq:frechet}
\end{equation}

\begin{equation}
\Phi_n = \sum_{o=1}^{N} T_o \left(w_o d_{no}\right)^{-\theta},
\qquad P_n = \gamma\, \Phi_n^{-1/\theta}
\label{eq:phi}
\end{equation}

\begin{equation}
\pi_{ni} = \frac{T_i \left(w_i d_{ni}\right)^{-\theta}}{\Phi_n}
\label{eq:share}
\end{equation}

\begin{equation}
w_i L_i = \sum_{n=1}^{N} \pi_{ni} X_n
\label{eq:clearing}
\end{equation}

\begin{equation}
X_n = w_n L_n + h(w) D_n, \qquad \sum_{n=1}^{N} D_n = 0
\label{eq:budget}
\end{equation}

\begin{equation}
\pi'_{ni}
= \frac{\hat{T}_i T^0_i \left(\hat{w}_i w^0_i \,\hat{d}_{ni} d^0_{ni}\right)^{-\theta}}
       {\sum_{o} \hat{T}_o T^0_o \left(\hat{w}_o w^0_o \,\hat{d}_{no} d^0_{no}\right)^{-\theta}}
\label{eq:pisub}
\end{equation}

\begin{equation}
\underbrace{\hat{T}_i\left(\hat{w}_i\hat{d}_{ni}\right)^{-\theta}}_{\text{hats}}
\cdot
\underbrace{T^0_i\left(w^0_i d^0_{ni}\right)^{-\theta}}_{=\ \Phi^0_n \pi^0_{ni}}
\label{eq:factor}
\end{equation}

\begin{equation}
\hat{\pi}_{ni}
= \frac{\pi'_{ni}}{\pi^0_{ni}}
= \frac{\hat{T}_i \left(\hat{d}_{ni}\hat{w}_i\right)^{-\theta}}
       {\sum_{o=1}^{N} \pi^0_{no}\, \hat{T}_o \left(\hat{d}_{no}\hat{w}_o\right)^{-\theta}}
\label{eq:pihat}
\end{equation}

\begin{equation}
\hat{\Phi}_n = \sum_{o=1}^{N} \pi^0_{no}\, \hat{T}_o
\left(\hat{d}_{no}\hat{w}_o\right)^{-\theta},
\qquad \hat{P}_n = \hat{\Phi}_n^{-1/\theta}
\label{eq:phihat}
\end{equation}

\begin{equation}
y^0_n \equiv \frac{w^0_n L^0_n}{X^0_n}
= \frac{\sum_{i} X^0_{in}}{\sum_{i} X^0_{ni}}
\label{eq:y0}
\end{equation}

\begin{equation}
h(w^0) D_n = \left(1 - y^0_n\right) X^0_n
\label{eq:transfer}
\end{equation}

\begin{align}
w'_n L'_n &= \hat{w}_n \hat{L}_n \, w^0_n L^0_n
           = \hat{w}_n \hat{L}_n \, y^0_n X^0_n \\[4pt]
h(w')D'_n &= h(w^0\hat{w})\,\hat{D}_n D_n
           = \frac{h(w^0\hat{w})}{h(w^0)}\,\hat{D}_n \cdot h(w^0)D_n
           = \frac{h(w^0\hat{w})}{h(w^0)}\,\hat{D}_n \left(1-y^0_n\right) X^0_n
\label{eq:terms}
\end{align}

\begin{equation}
\frac{X'_n}{X^0_n}
= y^0_n\, \hat{w}_n \hat{L}_n
+ \left(1-y^0_n\right) \frac{h(w^0\hat{w})}{h(w^0)}\, \hat{D}_n
\label{eq:xhat}
\end{equation}

\begin{equation}
\hat{w}_i \hat{L}_i = \sum_{n=1}^{N} \frac{\pi'_{ni} X'_n}{Y^0_i}
\label{eq:clearhat}
\end{equation}

\begin{equation}
\frac{\pi'_{ni} X'_n}{Y^0_i}
= \underbrace{\frac{\pi'_{ni}}{\pi^0_{ni}}}_{\text{Step 1}}
\cdot
\underbrace{\frac{\pi^0_{ni} X^0_n}{Y^0_i}}_{=\ X^0_{ni} / \sum_d X^0_{di}}
\cdot
\underbrace{\frac{X'_n}{X^0_n}}_{\text{Step 2}}
\label{eq:split}
\end{equation}

\begin{equation}
Z_i(\hat{w}) \equiv
\sum_{n=1}^{N}
\frac{X^0_{ni}}{\sum_{d=1}^{N} X^0_{di}}
\cdot
\frac{\hat{T}_i\left(\hat{d}_{ni}\hat{w}_i\right)^{-\theta}}
     {\sum_{o} \pi^0_{no}\hat{T}_o\left(\hat{d}_{no}\hat{w}_o\right)^{-\theta}}
\left(
 y^0_n \hat{w}_n \hat{L}_n
 + \left(1-y^0_n\right)\frac{h(w^0\hat{w})}{h(w^0)}\hat{D}_n
\right)
- \hat{w}_i \hat{L}_i
\label{eq:target}
\end{equation}

\begin{equation}
Z_i(\hat{w}) = 0 \qquad \text{for all } i
\label{eq:equil}
\end{equation}

## Question 2


```{r}
#download the data

country_list <- read_csv(here("data", "country_list.csv"))
trade <- read_csv(here("data", "bilateral_trade_country.csv"))

#set variable names
countries <- country_list$country
N         <- length(countries)

# trade elasticity, theta set to 4 as prescribed
theta <- 4

# building the matrix X0[n, i] = spending by n (destination) on goods from i (origin)
X0 <- matrix(0, nrow = N, ncol = N,
             dimnames = list(countries, countries))
X0[cbind(trade$country_dest, trade$country_org)] <- trade$trade_flow2014

# total spending of n = row sums; total income of i = column sums
spending <- rowSums(X0)
income   <- colSums(X0)

# expenditure shares pi0[n, i]
pi0 <- X0 / spending

# income-to-spending ratios
y0 <- income / spending

# share of i's sales going to n
sales_share <- t(t(X0) / income)
```

## Question 3


### (a) Initial guess

```{r q3a}
# shocks: 10% rise in Chinese productivity, nothing else
That <- rep(1, N)
names(That) <- countries
That["CHN"] <- 1.1

Dhat <- rep(1, N)   
Lhat <- rep(1, N)

# h(w) = w_US and what_US = 1, so the numeraire ratio is one
h_ratio <- 1

# step b = 0: guess what_i = 1 for all i
what <- rep(1, N)
names(what) <- countries

# US wage is the numeraire, fixed at 1 throughout
what["USA"] <- 1

what
```
### (b) Iteration

```{r q3b}
kappa <- 0.5
tol   <- 1e-10
B     <- 100000

# store the error at each step
err_path <- rep(NA, B)

what <- rep(1, N)
names(what) <- countries

for (b in 1:B) {

  Phihat <- as.vector(pi0 %*% (That * what^(-theta)))
  pihat  <- matrix(That * what^(-theta), nrow = N, ncol = N, byrow = TRUE) / Phihat
  Xhat   <- y0 * what * Lhat + (1 - y0) * h_ratio * Dhat

  Z <- colSums(sales_share * pihat * Xhat) - what * Lhat
  Z["USA"] <- 0

  err_path[b] <- max(abs(Z))

  if (err_path[b] < tol) break

  what <- what + kappa * Z
  what["USA"] <- 1
}

err_path <- err_path[1:b]

b
err_path[b]
```
### Convergence check

```{r}
# error falls monotonically at every step
all(diff(err_path) < 0)

# first few and last few errors
head(err_path)
tail(err_path)

# log scale makes the geometric decline visible
plot(err_path, type = "l", log = "y",
     xlab = "iteration b",
     ylab = "max |Z_i|  (log scale)",
     main = "Convergence of the excess demand error")
abline(h = tol, lty = 2)
```

## Question 4

### Real wage and real expenditure

```{r q4-results}
# change in market access and price index
Phihat <- as.vector(pi0 %*% (That * what^(-theta)))
Phat   <- Phihat^(-1 / theta)
names(Phihat) <- countries
names(Phat)   <- countries

# change in total spending
Xhat <- y0 * what * Lhat + (1 - y0) * h_ratio * Dhat

# change in domestic trade share
pihat <- matrix(That * what^(-theta), nrow = N, ncol = N, byrow = TRUE) / Phihat
dimnames(pihat) <- list(countries, countries)
pihat_ii <- diag(pihat)

# initial spending share on Chinese goods
pi0_china <- pi0[, "CHN"]

results <- data.frame(
  country   = countries,
  what      = what,
  Xhat      = Xhat,
  Phat      = Phat,
  real_wage = what / Phat,
  real_exp  = Xhat / Phat,
  pihat_ii  = pihat_ii,
  Phihat    = Phihat,
  pi0_china = pi0_china
)

results[order(-results$real_wage), ]
```


### Regressions

```{r q4-reg}
# drop China
reg_data <- subset(results, country != "CHN")

reg_data$ln_real_wage <- log(reg_data$real_wage)
reg_data$ln_real_exp  <- log(reg_data$real_exp)
reg_data$ln_pihat_ii  <- log(reg_data$pihat_ii)
reg_data$ln_Phihat    <- log(reg_data$Phihat)

# real wage on each regressor separately
summary(lm(ln_real_wage ~ ln_pihat_ii, data = reg_data))
summary(lm(ln_real_wage ~ ln_Phihat,   data = reg_data))
summary(lm(ln_real_wage ~ pi0_china,   data = reg_data))

# real expenditure on each regressor separately
summary(lm(ln_real_exp ~ ln_pihat_ii, data = reg_data))
summary(lm(ln_real_exp ~ ln_Phihat,   data = reg_data))
summary(lm(ln_real_exp ~ pi0_china,   data = reg_data))
```
The regression of the log real wage on $\ln\hat{\pi}_{ii}$ returns a coefficient
of exactly $-0.25$ with $R^2 = 1$. This is not an empirical finding but an
identity: for every country other than China $\hat{T}_i = 1$, so
$\hat{\pi}_{ii} = \hat{w}_i^{-\theta}/\hat{\Phi}_i$ and
$\hat{\pi}_{ii}^{-1/\theta} = \hat{w}_i/\hat{P}_i$. The estimate recovers
$-1/\theta$ because the model says it must. I take the exact fit as a check
that the solver is coded correctly.

Market access enters with a negative coefficient, which looks wrong until one
notices that $\hat{\Phi}_i$ is endogenous. Each country carries a large weight
in its own $\Phi$, so a country whose wage rises sees its own $\hat{\Phi}$ fall.
Taiwan has both the largest real wage gain and the lowest $\hat{\Phi}$. The
partial effect of market access is positive, but the regression picks up the
correlation with $\hat{w}_i$ and reverses the sign.

The initial Chinese import share gives the cleanest result: a coefficient of
0.0037 with an $R^2$ of 0.73. Countries that bought more from China to begin
with gain more, because the productivity improvement reaches them through
cheaper imports in proportion to how much of their spending goes to Chinese
goods. The real expenditure regressions behave the same way, except that the
coefficient on $\ln\hat{\pi}_{ii}$ is $-0.286$ rather than $-0.25$. The ACR
identity holds for real income, not real expenditure, and the gap between them
is the transfer.

## Question 5

### Solve with the Chinese wage as the transfer numeraire

```{r q5-solve}
kappa <- 0.5
tol   <- 1e-10
B     <- 100000

# store the error at each step
err_path2 <- rep(NA, B)

what2 <- rep(1, N)
names(what2) <- countries

for (b in 1:B) {

  Phihat2 <- as.vector(pi0 %*% (That * what2^(-theta)))
  pihat2  <- matrix(That * what2^(-theta), nrow = N, ncol = N, byrow = TRUE) / Phihat2

  # h(w0 what) / h(w0) = what_China
  h_ratio2 <- what2["CHN"]

  Xhat2 <- y0 * what2 * Lhat + (1 - y0) * h_ratio2 * Dhat

  Z2 <- colSums(sales_share * pihat2 * Xhat2) - what2 * Lhat
  Z2["USA"] <- 0

  err_path2[b] <- max(abs(Z2))

  if (err_path2[b] < tol) break

  what2 <- what2 + kappa * Z2
  what2["USA"] <- 1
}

err_path2 <- err_path2[1:b]

b
err_path2[b]
```

### Convergence check

```{r q5-convergence}
all(diff(err_path2) < 0)

head(err_path2)
tail(err_path2)

plot(err_path2, type = "l", log = "y",
     xlab = "iteration b",
     ylab = "max |Z_i|  (log scale)",
     main = "Convergence with the Chinese wage numeraire")
abline(h = tol, lty = 2)
```

### Real variables

```{r}
Phihat2 <- as.vector(pi0 %*% (That * what2^(-theta)))
Phat2   <- Phihat2^(-1 / theta)
names(Phat2) <- countries

h_ratio2 <- what2["CHN"]
Xhat2    <- y0 * what2 * Lhat + (1 - y0) * h_ratio2 * Dhat

results2 <- data.frame(
  country    = countries,
  what2      = what2,
  Phat2      = Phat2,
  real_wage2 = what2 / Phat2,
  real_exp2  = Xhat2 / Phat2
)

results2[order(-results2$real_wage2), ]
```

### Comparison with item 1.iii

```{r}
compare <- data.frame(
  country        = countries,
  y0             = y0,
  real_wage_US   = results$real_wage,
  real_wage_CHN  = results2$real_wage2,
  diff_wage      = results2$real_wage2 - results$real_wage,
  real_exp_US    = results$real_exp,
  real_exp_CHN   = results2$real_exp2,
  diff_exp       = results2$real_exp2 - results$real_exp
)

compare[order(-abs(compare$diff_wage)), ]

# the difference lines up with the sign of the trade imbalance
cor(compare$diff_wage, compare$y0)
plot(compare$y0, compare$diff_wage,
     xlab = "y0 (income / spending)",
     ylab = "change in real wage: China numeraire minus US numeraire")
abline(h = 0, lty = 2)
```

Switching the numeraire changes almost nothing for China but reshuffles
everyone else along a single dimension. The correlation between the change in
the real wage and $y^0_n$ is $-0.975$. The reason is that only one term in the
system differs: $(1-y^0_n)\,\hat{w}_{\text{China}}\,\hat{D}_n$. Since
$\hat{w}_{\text{China}} > 1$, every transfer is now worth more in real terms.
Deficit countries have $1 - y^0_n > 0$ and receive that transfer, so they spend
more and their wage rises. Surplus countries pay it and lose.

Ireland and the Netherlands run the largest surpluses in the sample
($y^0 = 1.11$ and $1.08$) and lose the most. Greece and the rest-of-world
aggregate run the largest deficits and gain the most. The effects on real
expenditure are roughly ten times larger than on real wages, up to 0.26
percentage points for Ireland, because the transfer enters spending directly
and reaches wages only through general equilibrium.

The magnitudes are small here because 2014 imbalances are modest, with $y^0$
ranging from 0.95 to 1.11. The point is that the choice of $h(w)$ is not a
harmless normalisation. It fixes the real value of international transfers and
therefore redistributes between debtors and creditors.